### 2021-09-14 tex/gen-sem.tex
[ writing ] various thesis additions

@@

If we want to capture a broader range of traversals, including not just
syntactic but also \emph{semantic} operations, we must be able to target things
other than terms, and act in an interesting way on term constructors.
Doing a straight generalisation of the type of \AgdaFunction{trav}, this
suggests that we want a function with the following type.

\missingfigure{Semantic traversal type signature}

>>>>

\def\SimpleSem{../agda/latex/SimpleSem.tex}

@@

If we want to capture a broader range of traversals, including not just
syntactic but also \emph{semantic} operations, we must be able to target things
other than terms, and act in an interesting way on term constructors.
Doing a straight generalisation of the type of \AgdaFunction{trav}, this
suggests that we want a function with the following type, where
\AgdaBound{$\C$} is the type family we are targeting.

\ExecuteMetaData[\SimpleSem]{semType}

Following the implementation of \AgdaFunction{trav}, we see that
\AgdaBound{$\C$} will need to support a semantic counterpart of each syntactic
form (\AgdaInductiveConstructor{var}, \AgdaInductiveConstructor{app}, and
\AgdaInductiveConstructor{lam}).
With syntactic kits, we already asked for the field \AgdaField{tm} to interpret
\AgdaBound{$\V$}-values as terms.
We rename \AgdaField{tm} to \AgdaField{⟦var⟧} to reflect its role in the
semantic traversal \AgdaFunction{sem}.
Now, we will also ask for fields to replace the right-hand side applications of
the other term constructors.
For application, we can stick with the obvious thing: we should be able to
combine a semantic function and its semantic argument to get the semantic
result.

\ExecuteMetaData[\SimpleSem]{semVarApp}

However, we want to treat binding constructs specially, particularly because
there are semanticses with no notion of binding.
We instead provide a function from values to computations that works in any
\emph{extension} of the current context.
Keeping \AgdaField{$\swarrow^k$} as before, we get the following semantic
replacement for \AgdaRecord{Kit}.

\ExecuteMetaData[\SimpleSem]{SemanticsExplicit}

With the aim of abstracting away from explicit contexts, bringing us closer to
natural deduction, we can use some new notation to rephrase these requirements.
We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
One of the basic connectives in this setting is the \emph{pointwise} arrow
\AgdaFunction{\_$\Rightarrow$\_}, which acts in
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
like the non-dependent arrow does in \AgdaPrimitive{Set}.
Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
to quantify over \emph{all} contexts.
Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
encapsulating the pattern of considering arbitrary \emph{extensions} of a
context.
To facilitate working in this point-free setting, I give infix versions of
the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
\AgdaFunction{\_$\V\vdash$\_} and \AgdaFunction{\_$\C\vdash$\_}).
The principal use of these aliases is to fill the right argument with a type
(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
a context given through the point-free machinery.

\ExecuteMetaData[\SimpleSem]{Circle}

\ExecuteMetaData[\SimpleSem]{SemanticsCircle}

To illustrate this definition, I will discuss a syntactic traversal ---
renaming --- and a semantic traversal --- a standard $\mathrm{Set}$ semantics.

For the renaming semantics, as with the renaming kit, we specify that
environments hold variables (\AgdaDatatype{\_$\ni$\_}) and show that variables
satisfy the required form of weakening (\AgdaFunction{$\swarrow^v$}).
Meanwhile, whereas all syntactic kits target terms
(\AgdaDatatype{\_$\vdash$\_}), with a semantic traversal we must specify the
target.
The fields \AgdaField{$\llbracket$var$\rrbracket$} and
\AgdaField{$\llbracket$app$\rrbracket$} follow straightforwardly, with variables
embedding into terms and a pair of terms of the right types giving an
application term in the same context, via the relevant constructors.
For the \AgdaField{$\llbracket$lam$\rrbracket$} case, we are given
\ExecuteMetaData[\SimpleSem]{bRenSemCircle}, and after producing a
\AgdaInductiveConstructor{lam}, are left needing a term in
\ExecuteMetaData[\SimpleSem]{resRenSemCircle}.
That the type of \AgdaBound{b} is wrapped in \AgdaFunction{$\bigcirc$} gives
us the ability to use \AgdaBound{b} in the extended context
\ExecuteMetaData[\SimpleKits]{GA}.
In particular, we point at the new variable to yield the desired term in the
same context.

\ExecuteMetaData[\SimpleSem]{RenSemCircle}

To produce a $\mathrm{Set}$ semantics, we shift from targeting terms to
targeting the interpretation of terms.
In particular, \ExecuteMetaData[\SimpleSem]{semGA}\ is the type of functions
from the interpretation of \AgdaBound{$\Gamma$} to the interpretation of
\AgdaBound{A}.
The interpretation of a type is defined as usual, by recursion on the structure
of the type.
The interpretation of a context is the interpretation for each of its types.
We still have environments storing variables, which delays the interpretation of
variables to the \AgdaField{$\llbracket$var$\rrbracket$} case and allows newly
bound variables to be referred to directly as variables, rather than fetching
them up-front from an environment of interpretations.
In the \AgdaField{$\llbracket$app$\rrbracket$} case, we have
\ExecuteMetaData[\SimpleSem]{mSetSemCircle}\ and
\ExecuteMetaData[\SimpleSem]{nSetSemCircle}, and combine them in the usual way
by distributing the interpretation of the context
\ExecuteMetaData[\SimpleSem]{gammaSetSemCircle}.
The \AgdaField{$\llbracket$lam$\rrbracket$} case involves the same placement of
the new variable into the environment as in \AgdaFunction{RenSem}.
Finally, we get the function \AgdaFunction{set} from terms to their
interpretations by passing in the identity $\ni$-environment \AgdaFunction{id}.

\ExecuteMetaData[\SimpleSem]{SetSemCircle}

The definition of \AgdaRecord{Semantics} above essentially enforces that the
term being traversed and the result of the traversal share the same binding
structure.
Concretely, \AgdaInductiveConstructor{lam} is the only case where we can bind
new variables, and at that point we must do exactly one binding.
This is fine for renaming and substitution, which preserve the binding
structure, and also for a standard denotational semantics, which is sufficiently
abstracted from binding.
However, if we want to do other syntactic translations --- for example,
converting from a syntax with $n$-ary functions to a syntax with only unary
Curried functions --- it would be useful to allow more choices when going under
a binder.
To this end, I replace the one-step binding modality \AgdaFunction{$\bigcirc$}
by the all-possible-renamings modality \AgdaFunction{$\Box$}.
\AgdaFunction{$\Box$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaBound{$\Gamma$}
states that \AgdaBound{T} holds not only at \AgdaBound{$\Gamma$}, but also at
any context \AgdaBound{$\Gamma^+$} containing \AgdaBound{$\Gamma$} (including
\ExecuteMetaData[\SimpleKits]{GD}\ for any \AgdaBound{$\Delta$}).

\ExecuteMetaData[\SimpleSem]{Box}

As well as the flexibility in binding structure, the \AgdaFunction{$\Box$}
modality allows us to use the somewhat more well behaved and standard
relation of renaming, rather than strict context extension.
The resulting definition of \AgdaRecord{Semantics} is as follows, and is
simply a version of the previous definition where \AgdaFunction{$\bigcirc$}
has been replaced by \AgdaFunction{$\Box$}.
It will become apparent when we implement the traversal \AgdaFunction{sem} why
the first field also changes to include a \AgdaFunction{$\Box$}.

\ExecuteMetaData[\SimpleSem]{Semantics}

Writing a \AgdaFunction{$\Box$}-based semantics is very similar to writing a
\AgdaFunction{$\bigcirc$}-based semantics, so I will only give one further
example.
I generalise the renaming example to derive a semantic traversal from any
syntactic traversal.
We need a slightly modified definition of \AgdaRecord{Kit} to provide
\emph{renaming} of \AgdaBound{$\V$}-values, rather than just extension.

\ExecuteMetaData[\SimpleSem]{Kit}

The interesting feature of the corresponding \AgdaRecord{Semantics} is that
we now pass \AgdaBound{b} the renaming \AgdaFunction{$\swarrow^v$}, projecting
the original context \AgdaBound{$\Gamma$} out of the
\AgdaInductiveConstructor{lam}-extended context
\ExecuteMetaData[\SimpleKits]{GA}.

\ExecuteMetaData[\SimpleSem]{kit}

### 2021-09-14 tex/lrkits.tex
[ writing ] various thesis additions

\begin{definition}[Usage-annotated recursive environment]\label{def:lr-rec-env}
  A \emph{recursive $\V$-environment} between annotated contexts $\Gamma$ and
  $\Delta$ is defined by cases on the shape of $\Delta$ (where
  $\Gamma \env\V_R \Delta$ is the notation for the
  type of recursive environments for given $\V$, $\Gamma$, and $\Delta$):
  \begin{itemize}
    \item There is one environment $\alr{} : \grP\gamma \env\V {\cdot}$
      whenever $\grP \leq \gr0$.
    \item For $\rho_l : \grPl\gamma \env\V_R \Delta_l$ and
      $\rho_r : \grPr\gamma \env\V \Delta_r$, we have an environment
      $\alr{\rho_l, \rho_r} : \grP\gamma \env\V_R \Delta_l, \Delta_r$ whenever
      $\grP \leq \grPl + \grPr$.
    \item For any value $v : \V\,\grPprime\gamma\,A$, we have an environment
      $\alr{v} : \grP\gamma \env\V_R \gr rA$ whenever
      $\grP \leq \gr r\grPprime$.
  \end{itemize}
\end{definition}

>>>>

\begin{definition}[Usage-annotated recursive environment]\label{def:lr-rec-env}
  A \emph{recursive $\V$-environment} between annotated contexts $\Gamma$ and
  $\Delta$ is defined by cases on the shape of $\Delta$ (where
  $\Gamma \env\V_R \Delta$ is the notation for the
  type of recursive environments for given $\V$, $\Gamma$, and $\Delta$):
  \begin{itemize}
    \item There is one environment $\alr{} : \grP\gamma \env\V {\cdot}$
      whenever $\grP \leq \gr0$.
    \item For $\rho_l : \grPl\gamma \env\V_R \Delta_l$ and
      $\rho_r : \grPr\gamma \env\V_R \Delta_r$, we have an environment
      $\alr{\rho_l, \rho_r} : \grP\gamma \env\V_R \Delta_l, \Delta_r$ whenever
      $\grP \leq \grPl + \grPr$.
    \item For any value $v : \V\,\grPprime\gamma\,A$, we have an environment
      $\alr{v} : \grP\gamma \env\V_R \gr rA$ whenever
      $\grP \leq \gr r\grPprime$.
  \end{itemize}
\end{definition}

### 2021-09-14 tex/part.tex
[ refactor ] put sec 5 in an \include
[ writing ] various thesis additions

Addition gives the structural rules.
Multiplication gives modalities.

Compare with other work.
About here, I can talk about Granule-style $\otimes$-pattern matching.

\subsection{Partiality (or not)} \todo{Appendix?}

As we have seen \todo{Have we?}, the way additive and multiplicative rules are
realised algebraically is related to models of separation logic.
Models of separation logic typically use \emph{partial} commutative monoids to
model a heap, so it is tempting to generalise the commutative monoid of
addition in our semirings to a \emph{partial} commutative monoid.
However, we find that the most natural notion of \emph{partial semiring} is
degenerate, in that all partial semirings are actually (total) semirings.

>>>>

As we have seen, the way additive and multiplicative rules are
realised algebraically is related to models of separation logic.
Models of separation logic typically use \emph{partial} commutative monoids to
model a heap, so it is tempting to generalise the commutative monoid of
addition in our semirings to a \emph{partial} commutative monoid.
However, we find that the most natural notion of \emph{partial semiring} is
degenerate, in that all partial semirings are actually (total) semirings.

### 2021-09-14 tex/semirings.tex
[ writing ] various thesis additions

\chapter{Usage restriction via semirings}

Addition gives the structural rules.
Multiplication gives modalities.

Compare with other work.
About here, I can talk about Granule-style $\otimes$-pattern matching.

\section{A usage-annotated calculus}\label{sec:lr}
\input{lr.tex}
\section{Linear natural deduction}\label{sec:lnd}
\input{lnd.tex}
\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Recursion}\label{sec:rec}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

>>>>

\chapter{Usage restriction via semirings}\label{sec:semirings}

The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
make crucial use of \emph{weakening} --- the fact that if we have
$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
We use this property to update environments as they go under binders.
However, as we saw in \cref{sec:linearity}, there are interesting calculi in
which general weakening does not hold.
As such, one of the aims of this chapter will be to find a form of weakening
applicable to variables of any type, while essentially retaining linearity
(as opposed to affinity).

As explained by McBride~\cite{McBride16}, the first insight is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether they are to be used or not.
We may use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(left) as a rule with usage annotations (right).
I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context is
now a list of types together with their usage annotations.
\todo{Reword: what's capital $\Gamma$}
If $\gamma = x_1 : A_1, \ldots, x_m : A_m$, then
$\gr r\gamma = \gr rx_1 : A_1, \ldots, \gr rx_m : A_m$.

\[
  \begin{prooftree}
    \hypo{\Gamma \vdash A}
    \hypo{\Delta \vdash B}
    \infer2{\Gamma, \Delta \vdash A \otimes B}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \hypo{\gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
  \end{prooftree}
\]

What the existence of $\gr0$ has bought us is that variables never go out of
scope in subterms; rather, we lose the ability to use certain variables.
Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
variables from $\delta$.
A rigorous proof of weakening for terms will come in \cref{sec:lrsub}.

If we follow the DILL style of variable management, there are not just the two
states \emph{used} ($\gr1$) and \emph{not used} ($\gr0$), but also
\emph{usable unrestrictedly}.
If we assign unrestricted (or \emph{intuitionistic}) variables an annotation
$\gr\omega$, we can make the following transcription of the DILL
$\otimes$-introduction rule.

\[
  \begin{prooftree}
    \hypo{\Theta; \Gamma \vdash A}
    \hypo{\Theta; \Delta \vdash B}
    \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
  \end{prooftree}
\]

To conceptualise the criteria on the usage annotations involved in this rule,
I introduce a formal $+$ operation on usage annotations.
The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
The capital calligraphic letters $\grP$, $\grQ$, and $\grR$ stand for all of
the usage annotations in a given context, with $\grR\gamma$ standing for
$\grR_1x_1 : \gamma_1, \ldots, \grR_mx_m : \gamma_m$.
Addition lifts pointwise to these vectors of usage annotations.
It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
rule on the right now accepts $\gr0$-annotated variables in its conclusion,
which is essential for weakening to be admissible.

\[
  \begin{prooftree}
    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \hypo{\grR = \grP + \grQ}
    \hypo{\grP\gamma \vdash A}
    \hypo{\grQ\gamma \vdash B}
    \infer3{\grR\gamma \vdash A \otimes B}
  \end{prooftree}
\]

Some other transcriptions:

\[
  \begin{prooftree}
    \infer0{\Theta; A \vdash A}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \hypo{\grR \leq \bra x}
    \hypo{\gamma_x = A}
    \infer2{\grR\gamma \vdash A}
  \end{prooftree}
\]

\[
  \begin{prooftree}
    \infer0{\Theta, A; {\cdot} \vdash A}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
  \end{prooftree}
  \quad\rightsquigarrow\quad
  \begin{prooftree}
    \hypo{\grR \leq \bra x}
    \hypo{\gamma_x = A}
    \infer2{\grR\gamma \vdash A}
  \end{prooftree}
\]

%In \cref{sec:simple}, we saw that the logical rules of simply typed
%$\lambda$-calculus can be described in terms of three basic premise combinators:
%$\dot1$, standing for no premises; $\dottimes$, allowing for multiple premises
%in the same context; and $\Theta \vdash A$, requiring a subterm of type $A$
%having bound the extra variables from context $\Theta$.
%However, we remember from \cref{sec:linearity} that in a substructural setting,
%we do not always want to copy assumptions for use in all subterms.
%This motivates me to introduce the additional premise combinators $I^*$, $\sep$,
%and $\cdot$ in \cref{sec:lnd}, allowing for the modes of usage exhibited in
%the introduction rules for $I$, $\otimes$, and $\oc$, respectively.

In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared previously in my work with Atkey~\cite{WA20}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\cite{Atkey18}.
The idea of \name{} is to augment simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
I use \cref{sec:lnd} to introduce some notation that allows us to restate the
typing rules of \name{} so as to not mention contexts explicitly.
The first problem I tackle with this new syntax is proving admissibility of
substitution.
I follow the methodology from \cref{sec:kits}, which

\section{A usage-annotated calculus}\label{sec:lr}
\input{lr.tex}
\section{Linear natural deduction}\label{sec:lnd}
\input{lnd.tex}
\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Adding recursion to \name{}}\label{sec:rec}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}
